---
title: "STA 6106 Homework 9"
author: "Tyler Grimes"
date: "April 11, 2016"
output: pdf_document
---

1. Suppose that the random variables $X$ and $Y$ both take on values in the interval $(0, B)$. Suppose that the conditional density of $X$ given that $Y = y$ is
$$f(x|y) = c(y)e^{-xy}, \ \ \ 0 < x < B$$
and the conditional density of $Y$ given that $X = x$ is
  $$f(y|x) = c(x)e^{-xy}, \ \ \ 0 < y < B$$
Give a method for simimately simulating the vector $X, Y$. Run a simulation to estimate (a) $E(X)$ and (b) $E(XY)$.  
  
# Algorithms
Since we are given conditional probabilities, Gibbs sampling can be used to generate observations from the marginal distributions. We will use Gibbs to generate values from $f(x)$, and then use $f(y | x)$ to generate a corresponding $y$. The following algorithm returns a random observation from the joint distribution of $X$ and $Y$. 
  
### Gibbs sampling algorithm
1.  Set $x_0$ to some initial value in (0, B), and set $i = 1$.
2.  Generate $y_i \sim f(y|x_{i - 1})$.
3.  Generate $x_i \sim f(x|y_{i})$.
4.  Repeat 2 and 3, $N$ times. 
5.  Return $(x, y) = (x_{N - 1}, y_N)$

A method for generating observations from the conditional distributions is required. The inverse-cdf method will be used. First we consider generating a random variable from $y \sim f(y|x)$. The cdf of $Y|X$ is
$$\begin{aligned}
F(y | x) &= \int_0^y f(y|x) dx \\
&= \int_0^y c(x)e^{-xy} dx \\
&= \frac{-c(x)}{x}\left( e^{-xy} - 1 \right), \ \ \ 0 < y < B
\end{aligned}$$
  
The inverse of this cdf is
$$\begin{aligned}
&u = \frac{-c(x)}{x}\left( e^{-xy} - 1 \right) \\
\Rightarrow &1 + \frac{-ux}{c(x)} = e^{-xy} \\
\Rightarrow &y = \frac{-1}{x}\ln\left( 1 - \frac{ux}{c(x)} \right)
\end{aligned}$$  
  
Note that $c(x)$ is just a constant factor that allows the pdf to integrate to $1$. Here, we have $\frac{1}{c(x)} = \int_0^B e^{-xy} dy = \frac{-1}{x} \left( e^{-By} - 1 \right)$. Assuming some value for $x$ and $B$ are given, the following algorithm can be used to generate an observation from the conditional distribution.  

### Inverse-CDF method for $Y|X$
1.  Generate a $u \sim U(0, 1)$
2.  Return $y = \frac{-1}{x}\ln \left( 1 - u\left( 1 - e^{-By} \right) \right)$.
  
And since the two conditional distributions have the same form, similar results are obtained for $X|Y$.
  
### Inverse-CDF method for $X|Y$
1.  Generate a $u \sim U(0, 1)$
2.  Return $x = \frac{-1}{y}\ln \left( 1 - u(1 - e^{-Bx}) \right)$.  
  
# Code
```{r}
#Generate a random observation from the conditional Y|X.
fy_conditional <- function(x, B = 10000) {
  u <- runif(1)
  -1/x * log(1 - u*(1 - exp(-B*x)))
}

#Generate a random observation from the conditional X|Y.
fx_conditional <- function(y, B = 10000) {
  u <- runif(1)
  -1/y * log(1 - u*(1 - exp(-B*y)))
}

#Generate an observation from the joint X, Y.
fx_and_y <- function(x0 = 1, B = 10000, BURN = 1000) {
  #Initialize x and compute the first value for y.
  x <- x0
  y <- fy_conditional(x, B)
  
  #Generate BURN (x, y) pairs for the burn in period.
  for(i in 1:BURN) {
    x <- fx_conditional(y, B)
    y <- fy_conditional(x, B)
  }
  
  #Use the last (x, y) pair as the observation.
  return(c(x, y))
}

#Simulate n observations from the joint distribution of X and Y.
n <- 10000
x <- vector("numeric", n)
y <- vector("numeric", n)

for(i in 1:n) {
  temp <- fx_and_y()
  x[i] <- temp[1]
  y[i] <- temp[2]
}

```

```{r echo = FALSE, fig.height = 6, fig.width = 12}
par(mfrow = c(1, 2))
plot(x, y)
plot(x, y, xlim = c(0, 6), ylim = c(0, 6))
```

We can estimate $E(X)$ by averaging the x values from our simulated observations.
```{r}
E_X <- mean(x)
```

This gives the approximation $E(X) \approx `r I(format(E_X, digits = 4))`$. Similarly, $E(XY)$ is estimated by averaging the products $x*y$ of each pair of observations from the simulated data.

```{r}
E_XY <- mean(x*y)
```

The approximation is $E(XY \approx `r I(format(E_XY, digits = 4))`$